---
title: "STATS 506 - Problem Set #1: Due September 12"
author: Sydney Yeomans
format: html
editor: visual
toc: true
---

## Problem 1

Download the data set about abalones (a type of sea snail) and do the following:

a. Import the data into a data.frame and give the appropriate column names.

```{r}
#| echo: true
abalones_df <- read.csv("abalone.data", header = FALSE)
head(abalones_df)
```
```{r}
#| echo: true
#Rename the columns based on what is in the abalone name file, names of variables are the following:
#V1 = Sex, V2 = Length, V3 = Diameter, V4 = Height, V5 = Whole weight, V6 = Shucked weight, V7 = Viscera weight, V8 = Shell weight, V9 = Rings

colnames(abalones_df) <- c("Sex", "Length", "Diameter", "Height", "Whole_weight", "Shucked_weight", "Viscera_weight", "Shell_weight", "Rings")

head(abalones_df)

```

b. The data contains information on three different sexes of abalone. Report the number of observations belonging to each sex.

```{r}
table(abalones_df$Sex)
```
In this data set there are 1307 females, 1528 males, and 1342 infants.

c.

i. 
```{r}
1 + 1
```

ii.
```{r}
1 + 1
```

iii.
```{r}
1 + 1
```

iv.
```{r}
1 + 1
```

d.

```{r}
1 + 1
```

e.

```{r}
1 + 1
```

## Problem 2

a. 

```{r}
1 + 1
```

b.

```{r}
2 * 2
```

c.

```{r}
1 + 1
```

d.

```{r}
1 + 1
```

e.

```{r}
1 + 1
```

f.

```{r}
1 + 1
```

h.

```{r}
1 + 1
```

## Problem 3

a.

```{r}
1 + 1
```

b.

```{r}
1 + 1
```

c.

```{r}
1 + 1
```
